---
title: "Data Science Specialization Capstone Project"
author: "Alejandro Osorio"
date: "September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set
library(tidyverse)
library(stringr)
```

## I. Appendix 1: Getting and Cleaning Data

### I.a. Reading Data

#### Twitter Dataset
```{r, echo = TRUE}
con <- file("en_US.twitter.txt", "r")
datatwitter <- data_frame(TEXT = readLines(con))
close(con)
```

Given the warning, the dimensions of both the obtained tibble and the original file, were compared.  They were equal.  Additionally, after some basic visual comparisons (head, tail, str, is.na), it was concluded that the obtained tibble was ok to work with.

#### Blogs Dataset
```{r, echo = TRUE}
con <- file("en_US.blogs.txt", "r")
datablogs <- data_frame(TEXT = readLines(con))
close(con)
```

After the same basic visual comparisons (head, tail, str, is.na) used with the previous dataset, it was concluded that the obtained tibble was also ok to work with.

#### News Dataset
```{r, echo = TRUE}
con <- file("en_US.news.txt", "r")
datanews_err <- data_frame(TEXT = readLines(con))
close(con)
```

Given the warning, the dimensions of both the obtained tibble and the original file, were compared.  The total number of lines obtained with the tibble (77,259) were much less than the ones from the original txt file (1,010,242).  Therefore, some cleaning was required, as follows.

### I.b. Cleaning Data

Further analysis to the "news" dataset, showed that line 77,259 was cut at the character "1", when compared to the one from the original file ("Ultra Edit" text editor was used for reading the original dataset, as well as the next cleaning steps). Therefore line 77,259 was analysed and an EOF character was found between texts "1" and "gallons", within the phrase "producing 1 -EOF- gallons".  Said character was deleted and the process re-run.  The result was a tibble with 766,277 rows (still less than the original file).  Another EOF character was found in row 766.277 (between texts "in" and "- inch", within the phrase "pears in EOF - inch slices."), deleted and the process re-run. Again, fewer rows were obtained (926,143).  Yet another EOF character was found in row 926,143 (between texts "in" and "-inch" within the phrase "Slice in EOF -inch-thick pieces"), deleted and the process re-run. Still, but last, fewer rows were obtained (948,564).  EOF character was found in row 926,143 (between texts "1" and "pounds" within the phrase "call for 1 EOF pounds of tomatoes"), deleted and the process re-run again.  Finally, 1,010,242 lines were obtained from the final fixed file, as follows:

```{r, echo = TRUE}
con <- file("en_US.news_FINAL.txt", "r")
datanews <- data_frame(TEXT = readLines(con))
close(con)
glimpse(datanews)
```

With this result, plus the same basic visual comparisons (head, tail, str, is.na) used with the previous datasets, it was concluded that the obtained tibble was also ok to work with and therefore the cleaning stage was considered finished.

### I.c. Preparing Preliminary Datasets

In order to work with sets words, each line of the obtained datasets was converted into a list of separate-underscore-words, as follows:

```{r, echo = TRUE}
words_twitter <- datatwitter$TEXT %>%
        str_to_lower() %>%
        str_extract_all(., boundary("word"))
words_blogs <- datablogs$TEXT %>%
        str_to_lower() %>%
        str_extract_all(., boundary("word"))
words_news <- datanews$TEXT %>%
        str_to_lower() %>%
        str_extract_all(., boundary("word"))
```

An example of the results obtained:

```{r}
head(words_twitter, 2)
```

## II, Appendix 2: Preliminary Analysis

### II.a. Basic Analysis Functions

#### Single words frequency counter

Frequency within a single row:

```{r}
rowordfrec <- function(dataset = prueba, rownum = 3) {
        uniquewordset <- unique(dataset[[rownum]])
        rowstats <<- data_frame()
        for(i in seq_along(uniquewordset)) {
                uniqueword <- paste("\\b", uniquewordset[i], "\\b", sep = "")
                uniquewordfrec <- str_count(dataset[rownum], uniqueword)
                rowstats[i,1] <<- uniquewordset[i]
                rowstats[i,2] <<- uniquewordfrec
                }
        names(rowstats) <<- c("word", "freq")
        }
```


### II.a. Twitter Dataset

Word Count
```{r, echo = TRUE}
str_count(datatwitter$TEXT, boundary("word"))
```

### II.b. Blogs Dataset

### II.c. News Dataset
